Cryptobenthic crab assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2025-07-03 manuscript.Rmd
fix cap figure confidence ellipses

```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone. Shallow samples are the same as those plotted with environmental drivers in @fig:fig_dbrda, to which the mesophotic samples (30 m, 60 m, and 90 m) are added here for comparison. Depth transect samples collected on the island of O&#x02BB;ahu are plotted with a diamond (and separately as Supplementary @fig:fig_supp_cap_oahu) for direct comparison relative to archipelago-wide shallow samples represented by a circle.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))
pcap <- pcap %>%
  mutate(
    oahu = factor(
      # if_else(island == "Oahu","oahu","noahu"),
      case_when(
        island == "Oahu" ~ "oahu",
        # island == "French Frigate Shoals" ~ "ffs",
        .default = "noahu"
      ),
      levels = c("oahu","ffs","noahu")
    )
  )

@@

newgrp <- "Shallow (O‘ahu)"
oldgrp <- 'Shallow (MHI)'
ll <- levels(pcap$sample_group)
after <- which(ll == oldgrp)

@@

# add the new factor level and make sure levels are appropriately ordered
# also, add a shape column that we'll use below
pcap <- pcap %>%
  mutate(
    sample_group = case_when(
      island == "Oahu" & depth_zone == "shallow" ~ newgrp,
      .default = sample_group
    ),
    sample_group = fct_reorder(sample_group,depth),
    shape = case_when(
      island == "Oahu" ~ 23,
      .default = 21
    )
  )

# get the new list of factor levels
depth_groups <- levels(pcap$sample_group)

# get a vector of shapes in the correct order
shapes <- pcap %>%
  distinct(sample_group,shape) %>%
  arrange(match(sample_group,depth_groups)) %>%
  deframe

# plot the cap
ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
  # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu"),guide="none") +
  scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
  guides(
    fill=guide_legend(override.aes=list(shape=shapes))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()

>>>>

```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone. Shallow samples are the same as those plotted with environmental drivers in @fig:fig_dbrda, to which the mesophotic samples (30 m, 60 m, and 90 m) are added here for comparison. Depth transect samples collected on the island of O&#x02BB;ahu are plotted with a diamond (and separately as Supplementary @fig:fig_supp_cap_oahu) for direct comparison relative to archipelago-wide shallow samples represented by a circle.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
pcap2 <- pcap %>%
  mutate(
    oahu = factor(
      # if_else(island == "Oahu","oahu","noahu"),
      case_when(
        island == "Oahu" ~ "oahu",
        # island == "French Frigate Shoals" ~ "ffs",
        .default = "noahu"
      ),
      levels = c("oahu","ffs","noahu")
    )
  )

centroids <- pcap2 %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))

@@

newgrp <- "Shallow (O‘ahu)"
oldgrp <- 'Shallow (MHI)'
ll <- levels(pcap2$sample_group)
after <- which(ll == oldgrp)

@@

# add the new factor level and make sure levels are appropriately ordered
# also, add a shape column that we'll use below
pcap2 <- pcap2 %>%
  mutate(
    old_sample_group = sample_group,
    sample_group = case_when(
      island == "Oahu" & depth_zone == "shallow" ~ newgrp,
      .default = sample_group
    ),
    dz=str_replace(depth_zone,"^([0-9]+)","z\\1"),
    dz=str_c(dz,region,depth)
  ) %>%
  arrange(dz) %>%
  mutate(
    sample_group = fct_inorder(sample_group),
    shape = case_when(
      island == "Oahu" ~ 23,
      .default = 21
    )
  )

# get the new list of factor levels
depth_groups <- levels(pcap2$sample_group)

# get a vector of shapes in the correct order
shapes <- pcap2 %>%
  distinct(sample_group,shape) %>%
  arrange(match(sample_group,depth_groups)) %>%
  deframe()

centroids <- pcap2 %>% group_by(old_sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2)) %>%
  mutate(old_sample_group = as.character(old_sample_group)) %>%
  rename(sample_group=old_sample_group)

# plot the cap
ggplot(pcap2,aes(x=LD1,y=LD2,fill=sample_group)) +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  stat_ellipse(mapping=aes(x=LD1,y=LD2,fill=old_sample_group),inherit.aes = FALSE,geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu"),guide="none") +
  scale_fill_manual(values=sample_depth_pal,name="Depth/region",guide="none") + 
  # geom_label_repel(data=centroids,aes(x=LD1,y=LD2,fill=sample_group,label=sample_group),inherit.aes = FALSE, color="white", show.legend = FALSE) +
  geom_label_repel(data=centroids,mapping=aes(x=LD1,LD2,label=sample_group,fill=sample_group),color="white",inherit.aes = FALSE,show.legend = FALSE) + 
  guides(
    fill=guide_legend(override.aes=list(shape=shapes))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()
